---
title: "Chapter 5 - Scale Effects on Terrain Attribute Calculation and Their Use as Environmental Covariates for Digital Soil Mapping"
author: "Stephen Roecker and James Thompson"
date: "2009"
output: 
  html_document: 
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r options, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r packages}
library(ggplot2)
library(dplyr)
```


# Case Study 2 - Soil and LSP correlations response to neighborhood size

## Tidy Soil Data

```{r soildata}

library(aqp)
library(dplyr)

fp <- "C:/Users/Stephen.Roecker/NextCloud/projects/thesis-wvu"

s <- read.csv(file.path(fp, "siteAttributes.csv"), stringsAsFactors = FALSE)
s[1:4] <- lapply(s[1:4], as.numeric)

h <- read.csv(file.path(fp, "horizonAttributes.csv"), stringsAsFactors = FALSE)
h$CaMg <- with(h, Ca + Mg) 
spc <- h
depths(spc) <- upedonid ~ hzdept + hzdepb

h_s <- aqp::slice(spc, 0:150 ~ fragvol + clay + sand + C + pH + CaMg, just.the.data = TRUE)
h_s$dep_int <- cut(h_s$hzdepb,
                   breaks = c(0, 50, 100, 150), 
                   labels = c("0-50", "50-100", "100-150")
                   )

h_di <- group_by(h_s, upedonid, dep_int) %>%
  summarize(fragvol = mean(fragvol, na.rm = TRUE),
            clay    = mean(clay,    na.rm = TRUE),
            C       = sum(C),
            pH      = mean(pH,      na.rm = TRUE),
            CaMg    = sum(CaMg), 
            hzthk   = sum(!is.na(.pctMissing))
            )%>%
  mutate(fragvol = ifelse(fragvol <= 0.1, 0.1, fragvol),
          C       = ifelse(C == 0,         NA,  C),
          CaMg    = ifelse(CaMg == 0,      NA,  CaMg)
          )
sh_di <- merge(s, h_di, by = "upedonid", all.x = TRUE)

```


## Depth Plots

```{r explore}

library(ggplot2)
library(GGally)

# Depth Plot

h_slab <- slab(spc, ~ fragvol + clay + C + pH + CaMg)

ggplot(h_slab, aes(x = bottom, y = p.q50)) +
  geom_line() +
  geom_ribbon(aes(ymin = p.q5,  ymax = p.q95, x = bottom), alpha = 0.2) + 
  geom_ribbon(aes(ymin = p.q25, ymax = p.q75, x = bottom), alpha = 0.2) + 
  xlim(150, 0) +
  facet_wrap(~ variable, scales = "free_x") +
  coord_flip() +
  xlab("depth (cm)") +
  ylab("5th, 25th, Median, 75th, and 95th Quantiles") +
  ggtitle("Depth Plot of Soil Properties")

```


## Scatter Plot Matrix

```{r spm}

vars <- c("fragvol", "clay", "C")
ggpairs(h[vars])
ggpairs(sh_di[vars])

```


## Create List of Variables

```{r data frame}

# construct comparison matrix
ns<- c(27, 45, 63, 81, 135, 189, 243)
gs <- c(9, 15, 27, 45, 81)

cm <- matrix(ns) %*% (1 / gs)
colnames(cm) <- gs
rownames(cm) <- ns

cm[cm != round(cm) & cm / 2 != round(cm / 2)] <- NA
cm[cm < 3]        <- NA
cm[cm %% 2 == 0]  <- NA
cm[upper.tri(cm)] <- NA
cm_df = data.frame(ns = row.names(cm), cm, check.names = FALSE)
knitr::kable(cm)


# construct data frame of geodata
geodata <- {
  expand.grid(source = "samb",
              res    = c("03m", "06m", "09m", "15m", "27m", "45m", "81m"),
              ws     = as.numeric(names(table(cm))),
              var    = c("", "slope", "aspect", "profcurv", "tancurv"),
              loc = "ug",
              format  = "tif",
              stringsAsFactors = FALSE
              ) ->.;
  # build file paths
  within(., {
    radius = (ws - 1) / 2
    tif = file.path("M:/geodata/project_data/dsm-utah-2009",
                   "upper_gauley",
                   paste0(source, res, "_", loc, 
                          ifelse(var != "", 
                                 paste0("_", var, "_", radius), 
                                 ""
                                 ),
                          ".tif")
                    )
    asc = file.path("M:/geodata/project_data/dsm-utah-2009",
                   "upper_gauley/asc",
                   paste0(source, res, "_", loc, 
                          ifelse(var != "", 
                                 paste0("_", var, "_", radius), 
                                 ""
                                 ),
                          ".asc")
                    )
    var = ifelse(var == "", "elev", var)
    }) ->.;
  subset(.,
         (res == "09m" & ws %in% cm_df$'9')  |
         (res == "15m" & ws %in% cm_df$'15') |
         (res == "27m" & ws %in% cm_df$'27') |
         (res == "45m" & ws %in% cm_df$'45') |
         (res == "81m" & ws %in% cm_df$'81')
         )->.;
  }


# convert the data frame to wide format
geodata_w <- reshape(geodata,
       direction = "wide",
       idvar     = c("res", "ws"),
       timevar   = "var",
       v.names   = c("tif", "asc")
       )

save(geodata, geodata_w, cm_df, file = "C:/Users/Stephen.Roecker/NextCloud/projects/dsm-utah-2009/geodata_df.RData")

```


## Sample Geodata and Correlate with Soil Depth Intervals

```{r geodata, eval = FALSE}

library(sp)
library(raster)

pts <- sh_di
coordinates(pts) <- ~ utm_easting + utm_northing
proj4string(pts) <- CRS("+init=epsg:26917")


# stack MuSTAnG rasters

load(file = "C:/workspace2/github/dsm-utah-2009/trunk/geodata_df.RData")

sg <- {
  subset(geodata, ! var == "elev") ->.;
  split(., .$res) ->.;
  lapply(., function(x) { 
    cat("stacking", x$res[1], "\n")
    rs = stack(x$asc)
    proj4string(rs) = "+init=epsg:26917"
    
    cat("extracting", x$res[1], "\n")
    sg = as.data.frame(raster::extract(rs, pts, sp = TRUE))
    
    # compute northness
    idx      = grepl("aspect", names(sg))
    sg[idx]  = lapply(sg[idx], function(x) abs(180 - x))
    
    return(sg)
    }) ->.;
  }

sg2 <- {
  lapply(sg, function(x) {
    cat("correlating and tidying", names(x)[16], "\n")
    # split by depth interval
    split(x, x$dep_int) ->.;
    # compute correlation and convert to long format
    lapply(., function(x2) {
      
      x2[x2 == 0] = NA
      x2 = na.exclude(x2)
      x2$fragvol = log(x2$fragvol)
      x2$C = log(x2$C)
      
      test = stack(data.frame(
        fragvol = cor(x2$fragvol, x2[16:ncol(x2)]), # , use = "pairwise.complete.obs"),
        clay    = cor(x2$clay,    x2[16:ncol(x2)]), # use = "pairwise.complete.obs"),
        C       = cor(x2$C,       x2[16:ncol(x2)]) # use = "pairwise.complete.obs"),
        ))
        test$ind = as.character(test$ind)
        # append depth interval
        test$dep_int = x2$dep_int[1]
        
        return(test)
        }) -> .;
    do.call("rbind", .) -> test;
    
    # split property.filename grouping variable into separate columns
    test = cbind(
      test,
      data.frame(
        do.call("rbind",
                strsplit(test$ind, "\\.|_")
                ),
        stringsAsFactors = FALSE
        )
      )
    names(test)[4:8] = c("prop", "source", "area", "var", "radius")
    
    # split source and variable columns into additional columns
    test = within(test, {
      res    = substr(source, 5, 8)
      res    = as.numeric(substr(res, 1, nchar(res) - 1))
      source = substr(source, 1, 4)
      radius = as.numeric(radius)
      ws     = radius + radius + 1
      ns     = ws * res
      res    = factor(res, levels = sort(unique(res)))
      dep_int = factor(paste(dep_int, "cm"), levels = c("0-50 cm", "50-100 cm", "100-150 cm"))
      var    = factor(var, levels = c("slope", "aspect", "profcurv", "tancurv"))
      })
    }) ->.;
  do.call("rbind", .) ->.;
  }

save(sg2, file = "C:/Users/Stephen.Roecker/NextCloud/projects/dsm-utah-2009/soil_correlations.RData")

```


## Create Line Plots

### 9-meter DEM and Neighborhood Size Combinations
```{r 9m-DEM, fig.dim = c(8, 10)}

load(file = "C:/Users/Stephen.Roecker/NextCloud/projects/dsm-utah-2009/soil_correlations.RData")

filter(sg2, res == "9") %>%
  ggplot(aes(x = ns, y = values, col = var, shape = res)) +
  geom_point(size = 2.5) +
  geom_line(lwd = 1) +
  geom_hline(yintercept = 0) +
  facet_grid(prop ~ dep_int) +
  theme(aspect.ratio = 1) +
  scale_shape_discrete(name = "grid size\n(meters)") +
  scale_color_discrete(name="terrain\nattribute") +
  ylim(- 0.5, 0.5) +
  xlab("neighborhood size (meters)") +
  ylab("correlation coefficient (r)") +
  labs(caption = "log transformation applied to fragvol") +
  ggtitle("Correlation between soil properties and terrain attributes\n over several neighborhood sizes")

```


### All DEM Grid Sizes and 3x3 Window Size Combinations
```{r 3x3, fig.dim = c(8, 10)}

filter(sg2, ws == 3) %>%
  ggplot(aes(x = ns, y = values, col = var)) +
  geom_point(aes(shape = res), size = 2.5) +
  geom_line(lwd = 1) +
  geom_hline(yintercept = 0) +
  facet_grid(prop ~ dep_int) +
  theme(aspect.ratio = 1) +
  scale_shape_discrete(name = "grid size\n(meters)") +
  scale_color_discrete(name="terrain\nattribute")+
  ylim(- 0.5, 0.5) +
  xlab("neighborhood size (meters)") +
  ylab("correlation coefficient (r)") +
  labs(caption = "log transformation applied to fragvol") +
  ggtitle("Correlation between soil properties and terrain attributes\n over several grid sizes")

```


### All DEM Grid and Neighborhood Size Combinations

```{r all-combinations, fig.dim = c(8, 10)}

ggplot(sg2, aes(x = ns, y = values, col = var, shape = res)) +
  geom_point(size = 2.5, alpha = 0.5) +
  geom_line(lwd = 1, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  facet_grid(prop ~ dep_int) +
  theme(aspect.ratio = 1) +
  scale_shape_discrete(name = "grid size\n(meters)") +
  scale_color_discrete(name="terrain\nattribute")+
  ylim(- 0.5, 0.5) +
  xlab("neighborhood size (meters)") +
  ylab("correlation coefficient (r)") +
  labs(caption = "log transformation applied to fragvol") +
  ggtitle("Correlation between soil properties and terrain attributes\n over several grid and neighborhood sizes")
```

